\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-04 paper.tex
Updated references

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress has led to a diversity of ARG definitions and representations.

We show that a synthesis of these views can be achieved by thinking
in terms of genomes and transmission
intervals.
% Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
% based on exhaustively detailing the ``events'' that occur
% in the history of a sample,
%are insufficient to describe the outputs of current methods.
We argue that classical ARG formulations, which assume
that all events can be known and precisely estimated,
are fundamentally unsuited to the realities
of modern, population-scale datasets.
We describe an alternative formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
Estimating the genetic genealogy of a set of DNA sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, encapsulating the resulting complex web of genetic
ancestry
(see \citet{lewanski2023era} for a biologically oriented introduction).
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into numerous downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,
guo2022recombination,
korfmann2022weak,
mahmoudi2022bayesian,
petr2022slendr,
rasmussen2022espalier,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}.
The technical details of \texttt{tskit}, and how it provides an
efficient and portable platform for ARG-based analysis, are beyond
the scope of this manuscript.
In the interest of avoiding confusion, however, we list a
few minor details in which the formal details of gARGs
provided in Section~\ref{sec-gARG} differ from their practical implementation in
\texttt{tskit}.
Firstly, ``edges'' in tree sequence terminology would perhaps be better
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes.
This denormalisation of the gARG data model is for efficiency purposes.
Secondly, zero- rather than one-based indexing is used for
nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
roots (rather than $0$ as used here for notational simplicity).

>>>>

\begin{abstract}
As a result of recombination, adjacent nucleotides can have
different paths of genetic inheritance and therefore
the genealogical trees for
% surely saying nucleotides and "sample of sequences" is concrete enough
% here? This paper isn't for a general audience, let's just say what
% we're talking about in direct language understandable by population
% geneticists.
a sample of DNA sequences vary along the genome.
The structure capturing the details of these intricately interwoven paths of
inheritance is referred to as an ancestral recombination graph (ARG).
New developments have made it possible to infer ARGs at scale,
enabling many new applications in population and statistical genetics.
This rapid progress has led to a diversity of ARG definitions and representations.
% We show that a synthesis of these views can be achieved by thinking
% in terms of genomes and transmission
% intervals.
% Standard mathematical formalisms,
% "detailing" here is intended to convey the we keep details about these
% events like recomb breakpoint.
% based on exhaustively detailing the ``events'' that occur
% in the history of a sample,
%are insufficient to describe the outputs of current methods.
We argue that classical ARG formulations, which assume
that all events can be known and precisely estimated,
are fundamentally unsuited to the realities
of modern, population-scale datasets.
We describe an alternative formulation that
encompasses the outputs of recent methods and
can capture the full richness of modern large-scale datasets.
By defining this ARG encoding in terms of
% we haven't define a "genome" yet, so need to give a bit of context
specific genomes and their intervals of genetic inheritance,
we avoid the need to exhaustively list (and estimate)
\emph{all} events. The effects of multiple events can be
aggregated in different ways, providing a natural way to express many
forms of approximate and partial knowledge about the
recombinant ancestry of a sample.
\end{abstract}

@@

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
% NOTE: slight inaccuracy of "DNA" here offset by making the problem concrete.
Estimating the genetic genealogy of a set of DNA sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, encapsulating the resulting complex web of genetic
ancestry
(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
introductions).
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards,deng2024robust}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood,
grundler2024geographic,
huang2024estimating}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into numerous downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,
guo2022recombination,
korfmann2022weak,
mahmoudi2022bayesian,
petr2023slendr,
rasmussen2022espalier,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood,
tsambos2023link,
tagami2024tstrait}.
The technical details of \texttt{tskit}, and how it provides an
efficient and portable platform for ARG-based analysis, are beyond
the scope of this manuscript.
In the interest of avoiding confusion, however, we list a
few minor details in which the formal details of gARGs
provided in Section~\ref{sec-gARG} differ from their practical implementation in
\texttt{tskit}.
Firstly, ``edges'' in tree sequence terminology would perhaps be better
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes.
This denormalisation of the gARG data model is for efficiency purposes.
Secondly, zero- rather than one-based indexing is used for
nodes in ARGs and oriented trees; consequently  $-1$ is used to denote the presence of
roots (rather than $0$ as used here for notational simplicity).
